Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-26 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
some more edits

There are two conditions that we must verify before we can confidently claim to
have placed a constraint on $\ell_{\rm th}$. First, as we have explained in Sec.~\ref{sec:review_theories}, all theories that we
consider must be interpreted as an EFT, meaning that it should be
considered valid only below an energy scale, or equivalently, a length-scale.
%
As a cut-off length-scale for the validity of the EFT we use,
%
\begin{equation}
\Lambda_{\rm EFT} (\varepsilon, \mathfrak{m}) = \varepsilon \, \frac{G \mathfrak{m}}{c^2} \,,
\label{eq:def_cutoff}
\end{equation}
%
where $\varepsilon$ is a dimensionless number and $\mathfrak{m}$ is the median
value of one of the mass scales involved in the problem.
%
We note that $\Lambda_{\rm EFT}$ has dimensions of length and hence can be compared to each theory's fundamental length-scale $\ell_{\rm th}$.
%
Here we explore the range $\varepsilon \in [0, 1]$, but following
Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr} we quote our final
results using $\varepsilon = 1/2$, \hs{but we stress that there is no
fundamental justification for this choice.}
% %
% In Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}, $\varepsilon = 1/2$ was used,
% but here we explore $\varepsilon \in [0, 1]$ for generality.
% %

@@

In Fig.~\ref{fig:dcs_cdf} we show the CDF for GW150914, we see that with $\gm = m_2$,
Eq.~\eqref{eq:eft_bound} is not satisfied unless $\varepsilon \approx 0.9$ (with only $j=0$ corrections) and
$\varepsilon \approx 0.7$ (with both $j=0$ and $1$ corrections).
%
The situation is different if we use $\gm = M_{\rm f}$. In this case, we find that
with or without spin corrections Eq.~\eqref{eq:eft_bound} can be satisfied
with $\varepsilon \leqslant 1/2$ (i.e., below the criteria used
Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}).
%
This means that with our model's assumptions and using the remnant's source mass $M_{\rm f}$ to set the
cut-off scale that we can claim an upper bound
%
\begin{equation}
\ell_{\rm dCS} \leqslant 42.6~\textrm{km}
\quad \textrm{at 90\% credible level},
\end{equation}
%
on dCS gravity and would constitute \emph{the first bound on this theory with GW
observations alone}.
%
We also note that since the CDF grows fast in approximate domain $\varepsilon = [0.2, 0.4]$,
that a stronger, albeit at lower credible level,
%
\begin{equation}
\ell_{\rm dCS} \leqslant 37.4~\textrm{km}
\quad \textrm{at 68\% credible level},
\end{equation}
%
can also be placed, when $N_{\rm max}=0$.

We can draw qualitatively similar conclusions from the GW200129 event.
In particular, we find,
%
\begin{equation}
\ell_{\rm dCS} \leqslant 36.6~\textrm{km}
% 43.9~\textrm{km}~( = \Lambda_{\rm EFT}(\tfrac{1}{2}, M_{f}) )
\quad \textrm{at 90\% credible level},
\end{equation}
%
and
%
\begin{equation}
\ell_{\rm dCS} \leqslant 29.1~\textrm{km}
% ~( = \Lambda_{\rm EFT}(0.3, M_{f}) )
\quad \textrm{at 68\% credible level},
\end{equation}
%
when $N_{\rm max}=0$.
%
These stronger bounds are a consequence of the larger support for $\ell_{\rm dCS} \lessapprox 15$~km
for GW200129 (compare the top and middle panels in Fig.~\ref{fig:dCS_exec_sum}), and in part due to
the smaller median remnant ($M_{\rm f} \approx 59.5\msun$ versus $M_{\rm f} \approx 61.8\msun$ for GW150914).
%
We also found for both GW events, that the perturbative-conditions~\eqref{eq:parspec_bound} required by the ParSpec
is violated for the $\gamma_{\rm dCS} \, \delta \tau^{(1)}$ coefficient.
%
This means that we cannot use this posterior to infer any meaningful bound on dCS gravity
and that is why we quoted only the $N_{\rm max}=0$ bound above.

@@

In Fig.~\ref{fig:cEFT_cdf} we show the CDF for both events, calculated in the
same way as already described for the EdGB and dCS theories.
%
We see that curves are very similar to those of dCS gravity for GW200129 (see
bottom panel in Fig.~\ref{fig:cEFT_cdf}).
%
Moreover, we find that the EFT~\eqref{eq:eft_bound} and
ParSpec~\eqref{eq:parspec_bound} bounds are satisfied for both events both when
$\mathfrak{m} = M_{\rm f}$ and $\varepsilon = 1/2$.
%
This allows us to place the combined bound of
%
\begin{equation}
    \ell_{\rm cEFT} \leqslant 38.2~\textrm{km}, \quad \textrm{at 90\% credible level}\,.
\end{equation}
%
As also happened for our study for dCS, the find that, for
the cubic EFT, the ParSpec bound is violated by the $N_{\rm max} = 1$ corrections to
the QNMs, meaning that we cannot use this case to draw any meaningful
constraint on this parameter.
%
We summarize our results in Table~\ref{tab:summary_ceft}.

@@

\begin{table}[t]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event & EFT    & ParSpec & Constraint ($\gm = M_{\rm f}$) \\
              &       & bound? & bound?  &                            \\
\hline
  & GW150914 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 51.7$~km \\
0 & GW200129 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 54.8$~km \\
  & Combined &     &      & \cellcolor{black!10}$\ell_{\rm qEFT} \leqslant 51.3$~km \\
% \hline
%   & GW150914 &  &     &    \\
% 1 & GW200129 &  &     &    \\
%   & Combined &  &     &    \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results the quartic EFT of GR for GW150914, GW200129, and
combined events using $\gm = M_{\rm f}$, $\varepsilon = 1/2$, and $N_{\rm max} = 0$. The quoted
result correspond \hs{to 90\% credible values \emph{if} we allow for a more flexible cut-off $\varepsilon \lesssim 0.65$.
However, the result is robust under the hard cut-off of $\varepsilon = 1/2$ at 65\% credible level.}
}
\label{tab:summary_qeft}
\end{table}

>>>>

There are two conditions that we must verify before we can confidently claim to
have placed a constraint on $\ell_{\rm th}$. First, as we have explained in Sec.~\ref{sec:review_theories}, all theories that we
consider must be interpreted as an EFT, meaning that they should be
considered valid only below an energy scale, or equivalently, a length-scale.
%
As a cut-off length-scale for the validity of the EFT we use,
%
\begin{equation}
\Lambda_{\rm EFT} (\varepsilon, \mathfrak{m}) = \varepsilon \, \frac{G \mathfrak{m}}{c^2} \,,
\label{eq:def_cutoff}
\end{equation}
%
where $\varepsilon$ is a dimensionless number and $\mathfrak{m}$ is the median
value of one of the mass scales involved in the problem.
%
We note that $\Lambda_{\rm EFT}$ has dimensions of length and hence can be compared to each theory's fundamental length-scale $\ell_{\rm th}$.
%
Here we explore the range $\varepsilon \in [0, 1]$, but following
Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr} we quote our final
results using $\varepsilon = 1/2$, \hs{but we stress that there is no
fundamental justification for this choice.}
% %
% In Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}, $\varepsilon = 1/2$ was used,
% but here we explore $\varepsilon \in [0, 1]$ for generality.
% %

@@

In Fig.~\ref{fig:dcs_cdf} we show the CDF for GW150914, we see that with $\gm = m_2$,
Eq.~\eqref{eq:eft_bound} is not satisfied unless $\varepsilon \approx 0.9$ (with only $j=0$ corrections) and
$\varepsilon \approx 0.7$ (with both $j=0$ and $1$ corrections).
%
The situation is different if we use $\gm = M_{\rm f}$. In this case, we find that
with or without spin corrections Eq.~\eqref{eq:eft_bound} can be satisfied
with $\varepsilon \leqslant 1/2$ (i.e., below the criteria used
Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}).
%
\hscomm{I commented out the 68\% results for consistency with the other sections.}
%
This means that with our model's assumptions and using the remnant's source mass $M_{\rm f}$ to set the
cut-off scale that we can claim an upper bound
%
\begin{equation}
\ell_{\rm dCS} \leqslant 41.9~\textrm{km}
\quad \textrm{at 90\% credible level},
\end{equation}
%
on dCS gravity and would constitute \emph{the first bound on this theory with GW
observations alone}.
%
% We also note that since the CDF grows fast in approximate domain $\varepsilon = [0.2, 0.4]$,
% that a stronger, albeit at lower credible level,
% %
% \begin{equation}
% \ell_{\rm dCS} \leqslant 37.4~\textrm{km}
% \quad \textrm{at 68\% credible level},
% \end{equation}
% %
% can also be placed, when $N_{\rm max}=0$.

We can draw qualitatively similar conclusions from the GW200129 event.
In particular, we find,
%
\begin{equation}
\ell_{\rm dCS} \leqslant 35.8~\textrm{km}
% 43.9~\textrm{km}~( = \Lambda_{\rm EFT}(\tfrac{1}{2}, M_{f}) )
\quad \textrm{at 90\% credible level}.
\end{equation}
% %
% and
% %
% \begin{equation}
% \ell_{\rm dCS} \leqslant 28.8~\textrm{km}
% % ~( = \Lambda_{\rm EFT}(0.3, M_{f}) )
% \quad \textrm{at 68\% credible level},
% \end{equation}
% %
% when $N_{\rm max}=0$.
% %
These stronger bounds are a consequence of the larger support for $\ell_{\rm dCS} \lessapprox 15$~km
for GW200129 (compare the top and middle panels in Fig.~\ref{fig:dCS_exec_sum}), and in part due to
the smaller median remnant ($M_{\rm f} \approx 59.5\msun$ versus $M_{\rm f} \approx 61.8\msun$ for GW150914).
%
We also found for both GW events, that the perturbative-conditions~\eqref{eq:parspec_bound} required by the ParSpec
is violated for the $\gamma_{\rm dCS} \, \delta \tau^{(1)}$ coefficient.
%
This means that we cannot use this posterior to infer any meaningful bound on dCS gravity
and that is why we quoted only the $N_{\rm max}=0$ bound above.

@@

In Fig.~\ref{fig:cEFT_cdf} we show the CDF for both events, calculated in the
same way as already described for the EdGB and dCS theories.
%
We see that curves are very similar to those of dCS gravity for GW200129 (see
bottom panel in Fig.~\ref{fig:cEFT_cdf}).
%
Moreover, we find that the EFT~\eqref{eq:eft_bound} and
ParSpec~\eqref{eq:parspec_bound} bounds are satisfied for both events both when
$\mathfrak{m} = M_{\rm f}$, $\varepsilon = 1/2$\hs{, and $N_{\rm max} = 0$.}
%
This allows us to place the combined bound of
%
\begin{equation}
    \ell_{\rm cEFT} \leqslant 38.2~\textrm{km}, \quad \textrm{at 90\% credible level}\,.
\end{equation}
%
As also happened for our study for dCS, the find that, for
the cubic EFT, the ParSpec bound is violated by the $N_{\rm max} = 1$ corrections to
the QNMs, meaning that we cannot use this case to draw any meaningful
constraint on this parameter.
%
We summarize our results in Table~\ref{tab:summary_ceft}.

@@

\begin{table}[h]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event & EFT    & ParSpec & Constraint ($\gm = M_{\rm f}$) \\
              &       & bound? & bound?  &                            \\
\hline
  & GW150914 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 51.7$~km \\
0 & GW200129 & Yes & Yes  & $\ell_{\rm qEFT} \leqslant 54.8$~km \\
  & Combined &     &      & \cellcolor{black!10}$\ell_{\rm qEFT} \leqslant 51.3$~km \\
% \hline
%   & GW150914 &  &     &    \\
% 1 & GW200129 &  &     &    \\
%   & Combined &  &     &    \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results the quartic EFT of GR for GW150914, GW200129, and
combined events using $\gm = M_{\rm f}$, $\varepsilon = 1/2$, and $N_{\rm max} = 0$. The quoted
result correspond \hs{to 90\% credible values \emph{if} we allow for a more flexible cut-off $\varepsilon \lesssim 0.65$.
However, the result is robust for the cut off $\varepsilon = 1/2$, at 65\% credible level.}
}
\label{tab:summary_qeft}
\end{table}
